Toward single particle reconstruction without particle picking: Breaking the detection limit

### 2018-06-11 TEX/draft4.tex
Implemented most comments of TB

\TODO{Revise--Cryo--electron microscopy (cryo--EM) is an innovative technology for single particle reconstruction (SPR) of macromolecules.} 
%In recent years, structures of many molecules, previously regarded as insurmountable, are now being reconstructed to near-atomic resolution
%; see for instance~\cite{kuhlbrandt2014resolution,bartesaghi20152,smith2014beyond}.
%This technological advancement was recognized by the 2017 Nobel Prize in Chemistry~\cite{nobel}. 
% 
In a cryo--EM experiment, biological samples are rapidly frozen in a thin layer of vitreous ice. Within the ice, the molecules are randomly oriented and positioned. The microscope produces a 2-D image of the samples embedded in the ice called a \emph{micrograph}. Each micrograph contains tomographic projections of the samples at unknown locations and under unknown viewing directions. The goal is to construct 3-D models of the molecules from the micrographs.

@@

%\paragraph{Particle picking.}
All contemporary methods in the field split the reconstruction procedure in several stages.
The first stage consists in extracting the various particle projections from the micrographs. This is called \emph{particle picking}. Later stages aim to construct a 3-D model of the molecule from these projections. The quality of the reconstruction eventually hinges on the quality of the particle picking stage. As can be seen from Figure~\ref{fig:micro_example}, locating the particles becomes increasingly challenging as the $\SNR$ degrades.

Crucially, it can be shown that reliable detection of individual particles is impossible below a certain critical $\SNR$. This fact has been recognized early on by the cryo-EM community. In particular, in a highly cited paper from 1995, Henderson~\cite{henderson1995limitations} investigates the following questions:
\begin{quote}
	\emph{For the purposes of this review, I would like to ask the question: what is the smallest size of free-standing molecule whose structure can in principle be determined by phase-contrast electron microscopy? Given what has already been demonstrated in published work, this reduces to the question: what is the smallest size of molecule for which it is possible to determine from images of unstained molecules the five parameters needed to define accurately its orientation (three parameters) and position (two parameters) so that averaging can be performed?}
\end{quote}
In that paper and in others that followed (e.g.,~\cite{glaeser1999electron}), it was established that particle picking is impossible for molecules below a certain weight.
%This has often been understood as meaning that particle reconstruction itself is impossible for such molecules~\cite{khoshouei2017cryo,liu2018nearatomic,liang2017phase}. \TODO{This list may be unfair, as some of these papers (in particular, \cite{liang2017phase}) just state that small molecules are very challenging, which is true.}
As a result, it is impossible to reconstruct such small molecules by any of the existing computational pipelines for single particle analysis, as the particles themselves cannot be picked from the micrographs.
%
%\TODO{either here, or in a close neighborhood of this paragraph you should mention the advances that the field has made in recent years towards reconstruction of small size molecules: mention Volta phase plate and the scaffolding cage (and cite the papers I sent you).}
This has motivated recent technical advances in the field, including the use of Volta phase plates~\cite{khoshouei2017cryo,liang2017phase} and scaffolding cages~\cite{liu2018nearatomic}.
%
%
Despite this progress, detecting small molecules (below 50 kDa) in the micrographs remains a challenge.
We note that nuclear magnetic resonance (NMR) spectroscopy and X-ray crystallography are well suited to reconstruct small molecules. Yet, cryo--EM has a lot to offer even for molecules with already known structures obtained via NMR spectroscopy or X-ray crystallography, because these methods have limited ability to distinguish conformational variability. \TODO{Need a ref for this claim.}

@@

Another interesting feature of the described approach pertains to model bias, whose importance in cryo-EM was stressed by a number of authors~\cite{vanheel1992correlation,shatsky2009method,henderson2013avoiding,vanheel2013finding}. In the classical ``Einstein from noise'' experiment, multiple realizations of pure noise are aligned to a picture of Einstein using cross-correlation and then averaged. In~\cite{shatsky2009method}, it was shown that the averaged noise rapidly becomes remarkably similar to the Einstein template. In the context of cryo--EM, this experiment exemplifies how prior assumptions about the particles may influence the reconstructed structure. This model bias is common to all particle picking methods based on template matching. In our approach, no templates are required, thus significantly reducing concerns about model bias.

@@

\begin{figure}[t]
	\centering
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_clean}
		\caption{$\sigma = 0$}
	\end{subfigure}%
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_s05}
		\caption{$\sigma = 0.5$}
	\end{subfigure}
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_s3}
		\caption{$\sigma = 3$}
	\end{subfigure}
	\caption{\label{fig:micro_example} Example of micrographs of size $250\times 250$ with additive white Gaussian noise of variance $\sigma^2$ for increasing values of $\sigma$. Each micrograph contain the same four occurrences of a $50 \times 50$ image of Einstein. In panel (c), the noise level is such that it is very challenging to locate the occurrences of the planted image. In fact, it can be shown that at low $\SNR$, reliable detection of individual image occurrences is impossible, even if the true image is known. By analogy to cryo-EM, this depicts a scenario where particle picking cannot be done.}	
\end{figure}

@@

%To support the claim that it may be possible to bypass particle picking in SPR from micrographs, as a proof of concept,
We conducted two experiments in the simplified image formation model described in the introduction:
\begin{enumerate}
	\item The first experiment aims to recover a 2-D image from an increasing number of micrographs with high noise, similar to the rightmost panel of Figure~\ref{fig:micro_example}. This is done using moments of second order, as these are sufficient to recover a 2-D image up to elementary symmetries;
	\item The second experiment aims to recover three distinct 1-D signals from an increasing number of 1-D micrographs with high noise. For this task, it is necessary to use moments up to third order.
\end{enumerate}
As outlined below, we find that it is indeed possible to recover good estimates of the ground truth signals from the highly corrupted micrographs, without particle picking. Furthermore, we find that the quality of estimation increases with the amount of data collected, despite the fact that particle picking remains challenging. The methods section provides additional details. In the discussion section, we outline how the general approach could be extended to full 3D SPR.

@@

We compute the average autocorrelation of the micrographs (equivalently, the average of their power spectra). This is a particularly simple computation. In the methods section, we show how, owing to separation of the occurrences, a determined portion of the averaged autocorrelation allows to estimate the power spectrum of the unknown image itself. Mathematically, it is easy to show that the quality of this estimate improves steadily as the amount of data grows, regardless of noise level. Then, to estimate the target image, we resort to a standard phase retrieval algorithm called relaxed-reflect-reflect (RRR)~\cite{bauschke2004rrr,elser2017rrr}. RRR is initialized with a picture of the physicist Isaac Newton and iterates to produce the estimate, up to a reflection ambiguity.\footnote{\TODO{Initializing with a face is debatable: it might help the algorithm. On the other hand, I agree that it supports the notion that we don't get trapped too easily.}}

Figure~\ref{fig:Einst_example} shows several estimated images for a growing number of micrographs. (A movie is available in \TODO{supplementary material}.) Figure~\ref{fig:error_per_micro} presents the normalized recovery error as a function of the amount of data available. Error is measured as the ratio of the root mean square error (RMSE) to the norm of the ground truth (square root of the sum of squared pixel intensities.) This is computed after fixing elementary symmetries (see methods.) As evidenced by these figures, the ground truth image can be estimated increasingly well from increasingly many micrographs, without particle picking.

@@

%\begin{figure}[h!]
%	\centering
%	\begin{subfigure}[h]{0.25\textwidth}
%		\centering
%		\includegraphics[scale=0.3]{Newton}
%		\caption{\label{fig:newton}Newton  (model)}
%	\end{subfigure}%
%	\begin{subfigure}[h]{0.25\textwidth}
%		\centering
%		\includegraphics[scale=0.3]{reconstruction2}
%		\caption{\small $P =1024$}
%	\end{subfigure}%
%	\begin{subfigure}[h]{0.25\textwidth}
%		\centering
%		\includegraphics[scale=0.3]{reconstruction20}
%		\caption{$P = 1024\times 10$}
%	\end{subfigure}%
%	\begin{subfigure}[h]{0.25\textwidth}
%		\centering
%		\includegraphics[scale=0.3]{reconstruction200}
%		\caption{$P = 1024\times 100$}
%	\end{subfigure}%
	\caption{\label{fig:Einst_example} Recovery of Einstein from micrographs at noise level $\sigma = 3$ (see Figure~\ref{fig:micro_example}(c)). Averaged autocorrelations of the micrographs allow to estimate the power spectrum of the target image. This does not require particle picking. A phase retrieval algorithm (RRR) produces the estimates here shown, initialized with an image of the physicist Isaac Newton. Estimates are obtained from $P$ micrographs (growing across panels), each containing $700$ image occurrences on average.---\TODO{Will add image with $P=512\times 10^4$ and image of Newton.}
		% At a noise level of $\sigma = 3$, this amounts to an $\SNR$ of $1/20$.
	}
\end{figure}

@@

In the second experiment, three 1-D signals, each of length $L = 21$, appear at random locations in one long 1-D signal, which we call a micrograph by analogy. Any two occurrences are separated by at least 20 entries. The signals appear respectively about 30, 20 and 10 million times in a micrograph of length 12.3 billion. The micrograph is then contaminated with additive white Gaussian noise. This results in an $\SNR$ of about $1/9$, while about 10\% of the micrograph contains signal. Neither the number of occurrences nor the noise level $\sigma$ are known.

In the methods section, we detail how autocorrelations of the micrograph can be used to estimate weighted averages of the autocorrelations of the target signals. The individual signals and their relative densities are then estimated from autocorrelations up to order three by solving a nonlinear least-squares problem.

Figure~\ref{fig:1Dheterosignals} shows how the estimates improve as we see a larger and larger fraction of the micrograph (that is, as more and more data becomes available.) As is clear from the picture, despite the high noise level which would make it very challenging to locate the signal occurrences, the individual signals can be estimated accurately given enough data. Furthermore, the propensity of each signal can also be estimated.

@@

\begin{figure}[t]
	\centering
	\includegraphics[width=\linewidth]{heterogeneous_progressive_n12300000000_466300}
	\caption{For the second experiment, each row shows, three times, one of the target signals (red), overlaid with an estimate (blue) obtained from a growing portion of the noisy micrograph (about $10^8$, $10^9$ and $10^{10}$ entries available to compute autocorrelations.) The last column depicts evolution of the relative root mean square error in estimating each individual target signal. Signals 1 to 3 appear respectively about 30.0, 20.0 and 10.0 million times. With the whole micrograph available, the algorithm estimated those to be 29.8, 21.9 and 10.0 million, respectively.
%		
%		An experiment for estimate three one-dimensional signals simultaneously with noise level $\sigma=3$. Red signals are the ground truth (targets) and the blue signals are our estimations. Individual RMSE of the estimates: $0.0239393 / 0.208925 / 0.0335956$. The estimated $\gamma$'s are $0.02574 / 0.01693 / 0.00818$ and the true ones $0.02561 / 0.01707 / 0.00853$. The $\SNR$ is $1/12$. \TODO{1. to replace with a ``progress'' plot 2. replace the triangle signal 3. add error progress figure}
	}
	\label{fig:1Dheterosignals}
\end{figure}

@@

The same general approach can, in principle, be applied directly to SPR from cryo-EM. Here, the micrographs contain numerous tomographic projections of molecules (possibly in different conformations) under unknown viewing directions. The aim is to estimate the 3-D volumes of the different conformations directly from micrographs. Each volume can be expanded linearly in a basis, so that the volume is characterized by its expansion coefficients. Since tomographic projection is a linear operation, autocorrelations of the micrographs (which can be estimated easily) are polynomial functions of the sought coefficients. Thus, autocorrelations of the micrographs provide a system of polynomial equations in the volume parameters, and the question becomes: are these equations sufficient to uniquely identify the volumes, and can we solve the system?

We show in Appendix~\ref{sec:cryoem} that the number of polynomial equations provided by the third-order autocorrelations is of the same order as the number of coefficients required to describe one volume at resolution comparable to that of the particle projections. This hints that it may be possible to reconstruct one or even several distinct volumes from these equations directly. Additional parameters in these equations could encode an unknown non-uniform viewing direction distribution and an unknown conformation distribution, which could then also be estimated. Crucially, the outlined approach involves no particle picking, hence a fortiori no viewing direction estimation or conformation clustering. As a result, it may not be limited to large molecules in the same way that particle picking approaches are. Concerns for model bias would also greatly be reduced.

@@

One possible concern is that the numerical experiments conducted here suggest a large amount of data may be necessary.\footnote{Whether or not this large amount of data would be necessary for any method to succeed given the unfavorable $\SNR$ is an interesting research question.} Recent trends in high-throughput cryo-EM technology \TODO{?} give hope that this may be a lesser concern in the long term. Still, large amounts of data also imply large amounts of computations. On this front, we note that computing autocorrelations of low orders can be done efficiently on CPUs and GPUs, and in parallel across micrographs. It can even be done in streaming mode, as only one look at each micrograph is necessary. The output of this data processing stage is a succinct summary in the form of autocorrelation estimates. Subsequent steps, which involve solving the system of polynomial equations, scale only in the size of that summary. Of course, an all important question then is whether the equations can be solved meaningfully in practice. The proof-of-concept experiments above suggest they might.

@@

Let $x_1,\ldots,x_K\in\RL$ be the sought signals and let $y\in\RN$ be the observed micrograph (notice that it is equivalent to think of the data as being one long micrograph or multiple shorter micrographs concatenated into one.)
The forward model (or ``image'' formation model) is as follows. For each target signal $x_k$, an unknown binary signal $s_k \in \{0, 1\}^{N-L+1}$ indicates (with 1's) the starting positions of all occurrences of $x_k$ in $y$, so that, with additive white Gaussian noise:
\begin{align}
	y & = \sum_{k=1}^K x_k \ast s_k + \varepsilon, & \varepsilon & \sim \mathcal{N}(0,\sigma^2 I_N),
	\label{eq:model}
\end{align}
where $\ast$ denotes (zero-padded) convolution. The binary signals obey the following property:
\begin{align}
	\textrm{If } s_k[i] = 1 \textrm{ and } s_{k'}[j] = 1 \textrm{ but } (k, i) \neq (k, j), \textrm{ then } |i - j| \geq 2L-1.
	\label{eq:spacing}
\end{align}
In words: the starting positions of any two occurrences (be it of the same signal or of two different signals) must be separated by at least $2L-1$ positions, so that their end points are necessarily separated by at least $L-1$ signal-free entries in the micrograph.

@@

Returning to the general case $K \geq 1$, let $M_1, \ldots, M_K$ denote the number of occurrences of signals $x_1, \ldots, x_K$ respectively, and define
\begin{align}
	\gamma_k & = \frac{M_k L}{N}, & \gamma & = \sum_{k=1}^K\gamma_k.
\end{align}
As above, we consider the asymptotic regime where $M_1,\ldots,M_K,N\to\infty$ while preserving the ratios $\gamma_k$ constant.
% 
Still under the spacing constraint~\eqref{eq:spacing}, similarly to the developments above, one can estimate a mixture of the autocorrelations of the $K$ target signals from the autocorrelations of the micrograph:
\begin{align}
\lim_{N\to\infty} a_y^1 & = \sum_{k=1}^K\gamma_k a_{x_k}^1, \nonumber\\
\lim_{N\to\infty} a_y^2[\ell] & = \sum_{k=1}^K\gamma_k a_{x_k}^2[\ell] +\sigma^2\delta[\ell],  \label{eq:data_ac}\\
\lim_{N\to\infty} a_y^3[\ell_1,\ell_2] & = \sum_{k=1}^K\gamma_k a_{x_k}^3[\ell_1,\ell_2] + \sigma^2\left(\sum_{k=1}^K\gamma_k a_{x_k}^1\right)(\delta[\ell_1,0]+\delta[0,\ell_2]+\delta[\ell_1,\ell_2]), \nonumber
\end{align}
where $\ell, \ell_1, \ell_2 = 0, \ldots, L-1$. The left hand side is straightforward to estimate from data: it provides a succinct summary of it. The right hand side involves polynomial functions of unknowns $\gamma_1, \ldots, \gamma_K, x_1, \ldots, x_K$. The task is to solve these polynomial equations in a robust way.

@@

For the 1-D experiment depicted in Figure~\ref{fig:1Dheterosignals}, we fix $K = 3$ signals of length $L = 21$. Following the forward model described at the beginning of this section, we generate an observation $y$ of length $12.3 \cdot 10^9$. Each of the three signals appears, respectively (and approximately) $30.0 \cdot 10^6$, $20.0 \cdot 10^6$ and $10.0 \cdot 10^6$ times in $y$ for a total of exactly $60 \cdot 10^6$ occurrences, such that at least $L-1$ zeros separate any two occurrences of any signals. 
This is done by randomly selecting $60 \cdot 10^6$ placements in $y$, one at a time with an accept/reject rule based on the separation constraint and locations picked so far. For each placement, one of the three signals is picked at random according to the proportions $1/2, 1/3, 1/6$. Then, i.i.d.\ Gaussian noise with mean zero and standard deviation $\sigma = 3$ is added, to form the observed $y$. The resulting $\SNR$ of $y$
% sqrt((m_want*sum(X.^2)')/(sigma^2*n))
is about 1/9.

@@

In this experiment, we assume we know the noise level $\sigma$ and the total number of occurrences of the target image across all micrographs.
In start contrast with the 1-D setup, the second-order autocorrelation determines almost any target image uniquely, up to reflection through the origin~\cite{hayes1982reconstruction,bendory2017fourier}.\footnote{\TODO{Should we cite \cite{bendory2017fourier} here? I'm not against citing it somewhere, but for the specific claim being supported here, it seems the older reference should be enough.}} This is because the second-order autocorrelations correspond to the Fourier magnitudes of the signal through the 2-D Fourier transform. 
Therefore, we estimate the signal's Fourier magnitudes (or power spectrum) from the Fourier magnitudes of the micrographs, at the cost of one 2-D fast Fourier transform (FFT) per micrograph. These can be computed highly efficiently and in parallel.

>>>>

\TODO{Revise--Cryo--electron microscopy (cryo--EM) is an innovative technology for single particle reconstruction (SPR) of macromolecules.} 
%In recent years, structures of many molecules, previously regarded as insurmountable, are now being reconstructed to near-atomic resolution
%; see for instance~\cite{kuhlbrandt2014resolution,bartesaghi20152,smith2014beyond}.
%This technological advancement was recognized by the 2017 Nobel Prize in Chemistry~\cite{nobel}. 
% 
In a cryo--EM experiment, biological samples are rapidly frozen in a thin layer of vitreous ice. Within the ice, the molecules are randomly oriented and positioned. The microscope produces a 2-D tomographic image of the samples embedded in the ice, called a \emph{micrograph}. Each micrograph contains tomographic projections of the samples at unknown locations and under unknown viewing directions. The goal is to construct 3-D models of the molecules from the micrographs.

@@

%\paragraph{Particle picking.}
All contemporary methods in the field split the reconstruction procedure into several stages.
The first stage consists in extracting the various particle projections from the micrographs. This is called \emph{particle picking}. Later stages aim to construct a 3-D model of the molecule from these projections. The quality of the reconstruction eventually hinges on the quality of the particle picking stage. Figure~\ref{fig:micro_example} illustrates how particle picking becomes increasingly challenging as the $\SNR$ degrades.

Crucially, it can be shown that reliable detection of individual particles is impossible below a certain critical $\SNR$. This fact has been recognized early on by the cryo-EM community. In particular, in an influential paper from 1995, Henderson~\cite{henderson1995limitations} investigates the following questions:
\begin{quote}
	\emph{For the purposes of this review, I would like to ask the question: what is the smallest size of free-standing molecule whose structure can in principle be determined by phase-contrast electron microscopy? Given what has already been demonstrated in published work, this reduces to the question: what is the smallest size of molecule for which it is possible to determine from images of unstained molecules the five parameters needed to define accurately its orientation (three parameters) and position (two parameters) so that averaging can be performed?}
\end{quote}
In that paper and in others that followed (e.g.,~\cite{glaeser1999electron}), it was established that particle picking is impossible for molecules below a certain weight.
%This has often been understood as meaning that particle reconstruction itself is impossible for such molecules~\cite{khoshouei2017cryo,liu2018nearatomic,liang2017phase}. \TODO{This list may be unfair, as some of these papers (in particular, \cite{liang2017phase}) just state that small molecules are very challenging, which is true.}
As a result, it is impossible to reconstruct such small molecules by any of the existing computational pipelines for single particle analysis in cryo--EM, as the particles themselves cannot be picked from the micrographs.
%
%\TODO{either here, or in a close neighborhood of this paragraph you should mention the advances that the field has made in recent years towards reconstruction of small size molecules: mention Volta phase plate and the scaffolding cage (and cite the papers I sent you).}
This has motivated recent technical advances in the field, including the use of Volta phase plates~\cite{khoshouei2017cryo,liang2017phase} and scaffolding cages~\cite{liu2018nearatomic}.
%
%
Despite this progress, detecting small molecules in the micrographs (below $\sim$50 kDa) remains a challenge.
We note that nuclear magnetic resonance (NMR) spectroscopy and X-ray crystallography are well suited to reconstruct small molecules. Yet, cryo--EM has a lot to offer even for molecules with already known structures obtained via NMR spectroscopy or X-ray crystallography, because these methods have limited ability to distinguish conformational variability. \TODO{Need a ref for this claim.}

@@

Another interesting feature of the described approach pertains to model bias, whose importance in cryo-EM was stressed by a number of authors~\cite{shatsky2009method,vanheel1992correlation,henderson2013avoiding,vanheel2013finding}. In the classical ``Einstein from noise'' experiment, multiple realizations of pure noise are aligned to a picture of Einstein using cross-correlation and then averaged. In~\cite{shatsky2009method}, it was shown that the averaged noise rapidly becomes remarkably similar to the Einstein template. In the context of cryo--EM, this experiment exemplifies how prior assumptions about the particles may influence the reconstructed structure. This model bias is common to all particle picking methods based on template matching. In our approach, no templates are required, thus significantly reducing concerns about model bias.

@@

\begin{figure}[t]
	\centering
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_clean}
		\caption{$\sigma = 0$}
	\end{subfigure}%
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_s05}
		\caption{$\sigma = 0.5$}
	\end{subfigure}
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_s3}
		\caption{$\sigma = 3$}
	\end{subfigure}
	\caption{\label{fig:micro_example} Example of micrographs of size $250\times 250$ with additive white Gaussian noise of variance $\sigma^2$ for increasing values of $\sigma$. Each micrograph contains the same four occurrences of a $50 \times 50$ image of Einstein. In panel (c), the noise level is such that it is very challenging to locate the occurrences of the planted image. In fact, it can be shown that at low $\SNR$, reliable detection of individual image occurrences is impossible, even if the true image is known. By analogy to cryo-EM, this depicts a scenario where particle picking cannot be done.}	
\end{figure}

@@

%To support the claim that it may be possible to bypass particle picking in SPR from micrographs, as a proof of concept,
We conducted two experiments in the simplified image formation model described in the introduction:
\begin{enumerate}
	\item The first experiment aims to recover a 2-D image from an increasing number of micrographs with high noise, similar to the rightmost panel of Figure~\ref{fig:micro_example}. This is done using moments of second order, as these are sufficient to recover a 2-D image up to elementary symmetries;
	\item The second experiment aims to recover three distinct 1-D signals from an increasing number of 1-D micrographs with high noise. For this task, it is necessary to use moments up to third order.
\end{enumerate}
As outlined below, we find that it is indeed possible to recover accurate estimates of the ground truth signals from the highly corrupted micrographs, without particle picking. Furthermore, we find that the quality of estimation increases with the amount of data collected, despite the fact that particle picking remains challenging. The Methods section provides additional details. In the discussion section, we outline how the general approach could be extended to full 3-D SPR.

@@

We compute the average autocorrelation of the micrographs (equivalently, the average of their power spectra). This is a particularly simple computation. In the methods section, we show how, owing to separation of the occurrences, a determined portion of the averaged autocorrelation allows to estimate the power spectrum of the unknown image itself. Mathematically, it is easy to show that the quality of this estimate improves steadily as the amount of data grows, regardless of noise level. Then, to estimate the target image, we resort to a standard phase retrieval algorithm called relaxed-reflect-reflect (RRR)~\cite{elser2017rrr}. % bauschke2004rrr,
RRR is initialized far away from the ground truth, and it iterates to produce the estimate, up to a reflection ambiguity.

Figure~\ref{fig:Einst_example} shows several estimated images for a growing number of micrographs, and a movie is available in \TODO{supplementary material}. Figure~\ref{fig:error_per_micro} presents the normalized recovery error as a function of the amount of data available. Error is measured as the ratio of the root mean square error (RMSE) to the norm of the ground truth (square root of the sum of squared pixel intensities.) This is computed after fixing elementary symmetries (see Methods.) As evidenced by these figures, the ground truth image can be estimated increasingly well from increasingly many micrographs, without particle picking.

@@

%\begin{figure}[h!]
%	\centering
%	\begin{subfigure}[h]{0.25\textwidth}
%		\centering
%		\includegraphics[scale=0.3]{Newton}
%		\caption{\label{fig:newton}Newton  (model)}
%	\end{subfigure}%
%	\begin{subfigure}[h]{0.25\textwidth}
%		\centering
%		\includegraphics[scale=0.3]{reconstruction2}
%		\caption{\small $P =1024$}
%	\end{subfigure}%
%	\begin{subfigure}[h]{0.25\textwidth}
%		\centering
%		\includegraphics[scale=0.3]{reconstruction20}
%		\caption{$P = 1024\times 10$}
%	\end{subfigure}%
%	\begin{subfigure}[h]{0.25\textwidth}
%		\centering
%		\includegraphics[scale=0.3]{reconstruction200}
%		\caption{$P = 1024\times 100$}
%	\end{subfigure}%
	\caption{\label{fig:Einst_example} Recovery of Einstein from micrographs at noise level $\sigma = 3$ (see Figure~\ref{fig:micro_example}(c)). Averaged autocorrelations of the micrographs allow to estimate the power spectrum of the target image. This does not require particle picking. A phase retrieval algorithm (RRR) produces the estimates here shown, initialized with an image of the physicist Isaac Newton. Estimates are obtained from $P$ micrographs (growing across panels), each containing $700$ image occurrences on average.
		\TODO{To add: initialization figure}
		% At a noise level of $\sigma = 3$, this amounts to an $\SNR$ of $1/20$.
	}
\end{figure}

@@

In the second experiment, three 1-D signals, each of length $L = 21$, appear at random locations in one long 1-D signal, which we call a micrograph by analogy. Any two occurrences are separated by at least 20 entries. The signals appear respectively about 30, 20 and 10 million times in a micrograph of length 12.3 billion. The micrograph is then contaminated with additive white Gaussian noise. This results in an $\SNR$ of about $1/9$, while about 10\% of the micrograph contains signal. Neither the number of occurrences nor the noise level $\sigma$ are known to the algorithm.

In the Methods section, we detail how autocorrelations of the micrograph can be used to estimate weighted averages of the autocorrelations of the target signals. The individual signals and their relative densities are then estimated from autocorrelations up to order three by solving a nonlinear least-squares problem.

Figure~\ref{fig:1Dheterosignals} shows how the estimates improve as we see a larger and larger fraction of the micrograph (that is, as more and more data becomes available.) As is clear from the picture, despite the high noise level which would make it very challenging to locate the individual signal occurrences, the signals can be estimated accurately given enough data. Furthermore, the \TODO{propensity\footnote{Might not be the right word; 'density' is not good because it might refer to the density of the particle for example, whereas here we mean to say the 'fraction of occurrences that come from a particular class'}} of each signal can also be estimated.

@@

\begin{figure}[t]
	\centering
	\includegraphics[width=\linewidth]{heterogeneous_progressive_n12300000000_466300}
	\caption{For the second experiment, each row shows, three times, one of the target signals (red), overlaid with an estimate (blue) obtained from a growing portion of the noisy micrograph (about $10^8$, $10^9$ and $10^{10}$ entries available to compute autocorrelations). The last column depicts evolution of the relative root mean square error in estimating each individual target signal. Signals 1 to 3 appear respectively about 30.0, 20.0 and 10.0 million times. With the whole micrograph available, the algorithm estimated those to be 29.8, 21.9 and 10.0 million, respectively.
%		
%		An experiment for estimate three one-dimensional signals simultaneously with noise level $\sigma=3$. Red signals are the ground truth (targets) and the blue signals are our estimations. Individual RMSE of the estimates: $0.0239393 / 0.208925 / 0.0335956$. The estimated $\gamma$'s are $0.02574 / 0.01693 / 0.00818$ and the true ones $0.02561 / 0.01707 / 0.00853$. The $\SNR$ is $1/12$. \TODO{1. to replace with a ``progress'' plot 2. replace the triangle signal 3. add error progress figure}
	}
	\label{fig:1Dheterosignals}
\end{figure}

@@

The same general approach can, in principle, be applied directly to SPR from cryo-EM. Here, the micrographs contain numerous tomographic projections of molecules (possibly in different conformations) taken from unknown viewing directions. The aim is to estimate the 3-D volumes of the different conformations directly from micrographs. Each volume can be expanded linearly in a basis, so that the volume is characterized by its expansion coefficients. Since tomographic projection is a linear operation, autocorrelations of the micrographs (which can be estimated easily) are polynomial functions of the sought coefficients. Thus, autocorrelations of the micrographs provide a system of polynomial equations in the volume parameters, and the question becomes: are these equations sufficient to uniquely identify the volumes, and can we solve the system?

We show in Appendix~\ref{sec:cryoem} that the number of polynomial equations provided by the third-order autocorrelations is of the same order as the number of coefficients required to describe one volume \TODO{at resolution comparable to that of the particle projections---remove?}. This hints that it may be possible to reconstruct one or even several distinct volumes from these equations directly. Additional parameters in these equations could encode an unknown viewing direction distribution and an unknown conformation distribution, which could then also be estimated. Crucially, the outlined approach involves no particle picking, hence a fortiori no viewing direction estimation or conformation clustering. As a result, it may not be limited to large molecules in the same way that particle picking approaches are. Concerns for model bias would also greatly be reduced.

@@

One possible concern is that the numerical experiments conducted here suggest a large amount of data may be necessary.\footnote{Whether or not this large amount of data would be necessary for any method to succeed given the unfavorable $\SNR$ is an interesting research question.} Recent trends in high-throughput cryo-EM technology \TODO{?} give hope that this may be a lesser concern in the long term. Still, large amounts of data also imply large amounts of computations. On this front, we note that computing autocorrelations of low orders can be done efficiently on CPUs and GPUs, and in parallel across micrographs. It can even be done in streaming mode, as only one look at each micrograph is necessary. The output of this data processing stage is a succinct summary in the form of autocorrelation estimates: its size is a function of the resolution, not a function of the number of observed micrographs. Subsequent steps, which involve solving the system of polynomial equations, scale only in the size of that summary. Of course, an important question then is whether the equations can be solved meaningfully in practice. The proof-of-concept experiments above suggest they might.

@@

Let $x_1,\ldots,x_K\in\RL$ be the sought signals and let $y\in\RN$ be the observed micrograph (notice that it is equivalent to think of the data as being one long micrograph or multiple shorter micrographs concatenated into one.)
The forward model (or ``image'' formation model) is as follows. For each target signal $x_k$, an unknown binary signal $s_k \in \{0, 1\}^{N-L+1}$ indicates (with 1's) the starting positions of all occurrences of $x_k$ in $y$, so that, with additive white Gaussian noise:
\begin{align}
	y & = \sum_{k=1}^K x_k \ast s_k + \varepsilon, & \varepsilon & \sim \mathcal{N}(0,\sigma^2 I_N),
	\label{eq:model}
\end{align}
where $\ast$ denotes linear convolution. The binary signals obey the following property:
\begin{align}
	\textrm{If } s_k[i] = 1 \textrm{ and } s_{k'}[j] = 1 \textrm{ but } (k, i) \neq (k', j), \textrm{ then } |i - j| \geq 2L-1.
	\label{eq:spacing}
\end{align}
In words: the starting positions of any two occurrences (be it of the same signal or of two different signals) must be separated by at least $2L-1$ positions, so that their end points are necessarily separated by at least $L-1$ signal-free entries in the micrograph.

@@

Returning to the general case $K \geq 1$, let $M_1, \ldots, M_K$ denote the number of occurrences of signals $x_1, \ldots, x_K$ respectively, and define
\begin{align}
	\gamma_k & = \frac{M_k L}{N}, & \gamma & = \sum_{k=1}^K\gamma_k.
\end{align}
As above, we consider the asymptotic regime where $M_1,\ldots,M_K,N\to\infty$ while preserving the ratios $\gamma_k$ constant.
% 
Still under the spacing constraint~\eqref{eq:spacing}, similarly to the developments above, one can estimate a mixture of the autocorrelations of the $K$ target signals from the autocorrelations of the micrograph:
\begin{align}
\lim_{N\to\infty} a_y^1 & = \sum_{k=1}^K\gamma_k a_{x_k}^1, \nonumber\\
\lim_{N\to\infty} a_y^2[\ell] & = \sum_{k=1}^K\gamma_k a_{x_k}^2[\ell] +\sigma^2\delta[\ell],  \label{eq:data_ac}\\
\lim_{N\to\infty} a_y^3[\ell_1,\ell_2] & = \sum_{k=1}^K\gamma_k a_{x_k}^3[\ell_1,\ell_2] + \sigma^2\left(\sum_{k=1}^K\gamma_k a_{x_k}^1\right)(\delta[\ell_1,0]+\delta[0,\ell_2]+\delta[\ell_1,\ell_2]), \nonumber
\end{align}
where $\ell, \ell_1, \ell_2 = 0, \ldots, L-1$. The left hand side is straightforward to estimate from data: it provides a succinct summary of it. The right hand side involves polynomial functions of unknowns $\gamma_1, \ldots, \gamma_K, x_1, \ldots, x_K$, and possibly $\sigma^2$. The task is to solve these polynomial equations in a robust way.

@@

For the 1-D experiment depicted in Figure~\ref{fig:1Dheterosignals}, we fix $K = 3$ signals of length $L = 21$. Following the forward model described at the beginning of this section, we generate an observation $y$ of length $12.3 \cdot 10^9$. Each of the three signals appears, respectively (and approximately), $30.0 \cdot 10^6$, $20.0 \cdot 10^6$ and $10.0 \cdot 10^6$ times in $y$ for a total of exactly $60 \cdot 10^6$ occurrences, such that at least $L-1$ zeros separate any two occurrences of any signals. 
This is done by randomly selecting $60 \cdot 10^6$ placements in $y$, one at a time with an accept/reject rule based on the separation constraint and locations picked so far. For each placement, one of the three signals is picked at random according to the proportions $1/2, 1/3, 1/6$. Then, i.i.d.\ Gaussian noise with mean zero and standard deviation $\sigma = 3$ is added, to form the observed $y$. The resulting $\SNR$ of $y$
% sqrt((m_want*sum(X.^2)')/(sigma^2*n))
is about 1/9.

@@

In this experiment, we assume we know the noise level $\sigma$ and the total number of occurrences of the target image across all micrographs.
In stark contrast with the 1-D setup, the second-order autocorrelation determines almost any target image uniquely, up to reflection through the origin~\cite{hayes1982reconstruction} (see also~\cite{bendory2017fourier} for a review). This is because the second-order autocorrelations correspond to the Fourier magnitudes of the signal through the 2-D Fourier transform. 
Therefore, we estimate the signal's Fourier magnitudes (or power spectrum) from the Fourier magnitudes of the micrographs, at the cost of one 2-D fast Fourier transform (FFT) per micrograph. These can be computed highly efficiently and in parallel.
